 

### 2022-05-31 thesis/methods.qmd
Finish thesis work

---
title: ''
format: pdf
crossref:
  sec-prefix: appendix
bibliography: ../references.bib
---

@@

The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker]. Any deviations from the preregistration are indicated in the 'Transparent Changes' document (@sec-tchanges). Major deviations will also be explicitly stated within the methods section. All data and code of this research project can be found on [GitHub](https://www.github.com/smnnlt/vo2max-processing).

I conducted all analyses using R Version 4.2.0 [@rcoreteam2022] in the R Studio IDE Version 2022.2.2.485 [@rstudioteam2022]. The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022]. The attached packages and default settings are documented in @sec-tech.

@@

The aim of the scoping review was to systematically map current practices of data processing for V̇O~2max~ determination in the scientific literature. Since determining V̇O~2max~ is a far too common procedure to perform an exhaustive search, I randomly sampled 500 articles that referred to V̇O~2max~ or similar keywords. Data on processing strategies were extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans. The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018]. @sec-checklist contains the corresponding reporting checklist.

@@

The article search was conducted on 16th March 2022 using [PubMed]() and [Web of Science](). The search included articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keywords. @tbl-search shows the exact search terms used.

@@

The abstracts from the articles included in the random sample were blinded for scanning. This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see @sec-blindabstract for an example). Two researchers independently scanned the abstracts to filter those that matched one of the exclusion criteria shown in @tbl-exclusion. When the screeners disagreed in their assessments, they resolved the conflict by discussion.

@@

-   metabolic cart used
-   measurement type (breath-by-breath, mixing chamber, ...)
-   type of outcome for V̇O~2max~ (primary, secondary, other)
-   data preprocessing (e.g., filtering)
-   data processing software
-   interpolation procedure
-   data processing/determination of V̇O~2max~:
    -   type (time average, breath average, digital filtering)
    -   alignment (rolling, binned, ...)
    -   interval (in seconds or breaths, parameters for filtering)
-   reference for the used data processing strategy

The criteria 'type of outcome' and 'reference' were added to the extraction list after the abstracts had been scanned, thus they were not stated in the preregistration. All extracted data is available as a csv file on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/data/extract/extraction.csv).

@@

excl_df <- data.frame(
    id = c("A*", "B*", "C*", "D*", "E", "F"),
    crit = c(
      "not in English", "no primary research", "research not in humans", "VO2max only estimated",
      "no appropriate test protocol", "no information regarding the exclusion criteria"),
    info = c(
      "Full text only available in non-English language",
      "research was no original investigation or only a reanalysis of data", 
      "research was conducted in animals", "VO2max was only approximated by means of a predictive equation", 
      "Protocol for VO2max testing did either not include exercise to voluntary exhaustion or was to long (>20 min) for a reliable estimate", 
      "no information regarding the exclusion criteria | crucial information on VO2max testing that allowed the evaluation of the other exclusion criteria were missing"
      )
)

@@

The extracted data is presented in a purely descriptive way. I calculated the relative and absolute frequency for the reporting of the extracted items. Similarly, I counted the use of different data strategies for processing data in all articles that reported measuring breath-by-breath. Total interval duration of averaging procedures were derived from the reported parameters.

@@

A total of N = 72 exercise tests were analysed for this study. Due to a miscalculation, the preregistration had incorrectly stated a number of 76 tests. The data were from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b]. The tested individuals were experienced distance runners (15 female, 54 male; three of the males participated in both studies). The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b], respectively, while using an identical exercise protocol. Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for 8 minutes at a velocity of 2.8 m·s^-1^ as a warm-up. After preparing the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m·s^-1^ for two minutes and subsequently increased velocity by 0.15 m·s^-1^ every 30 seconds. The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.

@@

The spiro Package version 0.0.4 for R [@nolte2022] processed the raw gas exchange data. It includes various algorithms to calculate V̇O~2max~ with user-defined parameters on given data. The full analysis script is available on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/comparison.md).

@@

Breath based moving averages were calculated on the raw data. As this functionality is implemented in the spiro package, I used the functions `spiro(data) |> spiro_max("30b")` for an exemplary 30-breath long averaging interval.

Butterworth filters are a class of recursive digital filters. Robergs et al. @robergs2010 and Weir et al. @weir2004 argue that these more advanced processing strategies are superior to the traditional moving or binned average approach for analysing gas exchange data. However Robergs et al. @robergs2010 missed to account for the time lag introduced by Butterworth filters. Therefore I applied a zero-phase Butterworth filter by means of forward-backward filtering. While this effectively zeroes out the time lag, the resulting filter is non-causal, which means it cannot be used online (i.e. in real time). However, for the present application, an offline filter is sufficient. The forward-backward filtering also introduces transients at both ends of the signal [@gustafsson1996]. I therefore padded the reverted signal at both the start and the end of the array (identical to the 'even padtype' in Python's `scipy.signal.filtfilt()` function [@virtanen2020]). I used 3 as the order parameter and 0.04 as the low-pass cut-off frequency for each filter [@robergs2010]. Note that due to the forward-backward approach both the filter's overall order and overall cut-off frequency are different from these values. The described approach is implemented in the spiro package [@nolte2022], and can be used by calling, for example, the functions `spiro(data) |> spiro_max("0.04fz3")`.

### Comparison of methods

To compare different processing methods within and between individuals, I chose to express the V̇O~2max~ normalized to a reference procedure. The reference procedure was chosen as being the most commonly applied in current literature as determined by the systematic review. Individual V̇O~2max~ values were expressed in reference to this procedure, where a value of 1 means that the processing method yields exactly the same V̇O~2max~ value as the reference method. I calculated the data for all integer parameter values within the range of the values found in the literature during the review. On a group level, I calculated the median and 10%- and 90%-quantiles of each processing strategy.

>>>>

---
title: ''
format: pdf
bibliography: ../references.bib
---

@@

The work presented in this thesis was preregistered before the start of the project on the [Open Science Framework](https://osf.io/3am4s) [@foster2017], following the 'Inclusive Systematic Review Registration Form' [@akker]. Any deviations from the preregistration are indicated in the 'Transparent Changes' document ([Appendix @sec-tchanges]). Major deviations will also be explicitly stated within the methods section. All data and code of this research project can be found on [GitHub](https://www.github.com/smnnlt/vo2max-processing).

I conducted all analyses using R Version 4.2.0 [@rcoreteam2022] in the R Studio IDE Version 2022.2.2.485 [@rstudioteam2022]. The thesis was entirely written in Quarto Version 0.9.380 [@allaire2022]. The attached packages and default settings are documented in [Appendix @sec-tech].

@@

The aim of the scoping review was to systematically map current practices of data processing for V̇O~2max~ determination in the scientific literature. Since determining V̇O~2max~ is a far too common procedure to perform an exhaustive search, I randomly sampled 500 articles that referred to V̇O~2max~ or similar keywords. Data on processing strategies were extracted from all sampled articles that directly measured V̇O~2max~ using an appropriate testing procedure in humans. The review was performed in accordance to the PRISMA extension for Scoping reviews [@tricco2018]. [Appendix @sec-checklist] contains the corresponding reporting checklist.

@@

The article search was conducted on 16th March 2022 using [PubMed](https://pubmed.ncbi.nlm.nih.gov/) and [Web of Science](https://www.webofscience.com/wos/woscc/basic-search). The search included articles published from 2017 to 2022 referring to 'maximum oxygen uptake' or equivalent terms in title, abstract or keywords. @tbl-search shows the exact search terms used.

@@

The abstracts from the articles included in the random sample were blinded for scanning. This meant removing any further information not relevant for the screening---such as authors or journal---leaving only the title, abstract and an ID of the article (see [Appendix @sec-blindabstract] for an example). Two researchers independently scanned the blinded abstracts to filter those that matched one of the exclusion criteria shown in @tbl-exclusion. When the screeners disagreed in their assessment, they resolved the conflict by discussion.

@@

-   metabolic cart used
-   measurement type (breath-by-breath, mixing chamber)
-   type of outcome for V̇O~2max~ (primary, secondary, other)
-   data preprocessing (e.g., filtering)
-   data processing software
-   interpolation procedure
-   data processing/determination of V̇O~2max~:
    -   type (time average, breath average, digital filtering, ...)
    -   alignment (moving, binned, ...)
    -   interval (in seconds or breaths, parameters for filtering)
-   rationale for the used data processing strategy (e.g. a reference)

The criteria 'type of outcome' and 'rationale' were added to the extraction list after the abstracts had been scanned, thus they were not stated in the preregistration. All extracted data is available as a csv file on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/data/extract/extraction.csv).

@@

excl_df <- data.frame(
    id = c("A*", "B*", "C*", "D*", "E", "F"),
    crit = c(
      "not in English", "no primary research", "research not in humans", "VO2max only estimated",
      "no appropriate test protocol", "no information regarding the exclusion criteria"),
    info = c(
      "full text only available in non-English language",
      "research was no original investigation or only a reanalysis of data", 
      "research was conducted in animals", "VO2max was only approximated by means of a predictive equation", 
      "protocol for VO2max testing did either not include exercise to voluntary exhaustion or was to long (>20 min) for a reliable estimate", 
      "crucial information on VO2max testing that allowed the evaluation of the other exclusion criteria were missing"
      )
)

@@

The extracted data is presented in a purely descriptive way. I calculated the relative and absolute frequency for the reporting of the extracted items. Similarly, I counted the use of different strategies for processing data in all articles that reportedly measured breath-by-breath. Total interval duration of averaging procedures were derived from the reported parameters.

@@

A total of N = 72 exercise tests were analyzed for this study. Due to a miscalculation, the preregistration had incorrectly stated a number of 76 tests. The data were from previous research on the metabolic profile of endurance runners [@quittmann_unpubl_a; @quittmann_unpubl_b]. The tested individuals were experienced distance runners (15 female, 54 male; three of the males participated in both studies). The V̇O~2max~ tests were conducted in March to September 2019 [@quittmann_unpubl_a] and March to October 2021 [@quittmann_unpubl_b], respectively, while using an identical exercise protocol. Participants run on a treadmill (saturn 300/100, h/p/cosmos sports & medical 127 GmbH, Nussdorf-Traunstein, Germany) with 1% inclination for 8 minutes at a velocity of 2.8 m·s^-1^ as a warm-up. After preparing the gas exchange measures, they started a ramp protocol with an initial speed of 2.8 m·s^-1^ for two minutes and subsequently increased velocity by 0.15 m·s^-1^ every 30 seconds. The researchers provided verbal encouragement and terminated the exercise when the participants reached subjective exhaustion.

@@

The spiro package version 0.0.4 for R [@nolte2022] processed the raw gas exchange data. It includes various algorithms to calculate V̇O~2max~ with user-defined parameters on given data. The full analysis script is available on [GitHub](https://github.com/smnnlt/vo2max-processing/blob/master/scripts/comparison.md).

@@

Breath based moving averages were calculated on the raw data. As this functionality is implemented in the spiro package [@nolte2022], I used the functions `spiro(data) |> spiro_max("30b")` for an exemplary 30-breath long averaging interval.

Butterworth filters are a class of recursive digital filters. Robergs et al. @robergs2010 and Weir et al. @weir2004 argued that these more advanced processing strategies are superior to the traditional moving or binned average approach for analysing gas exchange data. However Robergs et al. @robergs2010 missed to account for the time lag introduced by Butterworth filters. Therefore I applied a zero-phase Butterworth filter by means of forward-backward filtering. While this effectively zeroes out the time lag, the resulting filter is non-causal, which means it cannot be used online (i.e. in real time). However, for the present application, an offline filter is sufficient. The forward-backward filtering also introduces transients at both ends of the signal [@gustafsson1996]. I therefore padded the reverted signal at both the start and the end of the array (identical to the 'even padtype' in Python's `scipy.signal.filtfilt()` function [@virtanen2020]). I used 3 as the order parameter and 0.04 as the low-pass cut-off frequency for each filter [@robergs2010]. Note that due to the forward-backward approach both the filter's overall order and overall cut-off frequency are different from these values. The described approach is implemented in the spiro package [@nolte2022], and can be used by calling, for example, the functions `spiro(data) |> spiro_max("0.04fz3")`.

### Comparison of Methods

To compare different processing methods within and between individuals, I chose to express the V̇O~2max~ normalized to a reference procedure. The reference procedure was chosen as being the most commonly applied in current literature as determined by the scoping review. Individual V̇O~2max~ values were expressed in reference to this procedure, where a value of 1 means that the processing method yields exactly the same V̇O~2max~ value as the reference method. I calculated the data for all integer parameter values within the range of the values found in the literature during the review. On a group level, I calculated the median and 10%- and 90%-quantiles of each processing strategy.

### 2022-05-31 thesis/results.qmd
Finish thesis work

```{r}
#| echo: false
#| fig-cap: "Flow diagram for the systematic scoping review in accordance with the PRISMA 2020 Statement [@page2021]"
#| fig-scap: "Flow diagram for the systematic scoping review"
#| label: fig-flow
#| out-width: '90%'

@@

```{r}
#| echo: false
#| fig-cap: "Data strategies for processing breath-by-breath data in the reviewed literature (n = 88)."
#| label: fig-strategies
#| out-width: '90%'

@@

The calculation intervals for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration). 30 second length intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.

@@

```{r}
#| echo: false
#| fig-cap: "V̇O~2max~ varies by data processing strategy. Values are expressed relative to the V̇O~2max~ from a 30-second binned average — the most common strategy as determined by the review. Solid lines display the median, the shaded area marks the interval between 10th and 90th percentile. Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages. Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%."
#| fig-scap: "Comparison of different data processing strategies"
#| label: fig-comparison
#| out-width: '90%'

@@

Moving time and moving breath averages yield nearly identical values for V̇O~2max~ over all calculation intervals. This corresponds to the trained athletes reaching respiratory rates around 60 min^-1^ in the final minute of the exercise test (see @fig-rr). Using a zero-phase forward-backward filter (third order, 0.04 Hz cut-off frequency) lead to V̇O~2max~ values 0.4% \[-0.2%; +1.4%\] (median, 10th and 90th quantile) higher than the 30-second binned time average approach.

```{r}
#| echo: false
#| fig-cap: "Respiratory rates peak around 60 min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72)."
#| fig-scap: "Respiratory rates during the ramp tests"
#| label: fig-rr
#| out-width: '90%'

>>>>

```{r}
#| echo: false
#| fig-cap: "Flow diagram for the systematic scoping review in accordance with the PRISMA 2020 Statement [@page2021]"
#| fig-scap: "Flow diagram for the systematic scoping review."
#| label: fig-flow
#| out-width: '90%'

@@

```{r}
#| echo: false
#| fig-cap: "Strategies for processing breath-by-breath data in the reviewed literature (n = 88)."
#| label: fig-strategies
#| out-width: '90%'

@@

The calculation intervals for time-based averages of mixing chamber and breath-by-breath devices ranged from 5 to 60 seconds (see @fig-duration). 30-second intervals were most common to define V̇O~2max~, while authors also often employed shorter (10-20 s) and longer (60 s) periods.

@@

```{r}
#| echo: false
#| fig-cap: "V̇O~2max~ varies by data processing strategy. Values are expressed relative to the V̇O~2max~ from a 30-second binned average — the most common strategy as determined by the review. Solid lines display the median, the shaded areas mark the interval between 10th and 90th percentile. Using moving average leads to systematically higher V̇O~2max~ values compared to binned time averages. Changing the averaging interval (in seconds or breaths) can lead to median changes in V̇O~2max~ as large as 5%."
#| fig-scap: "Comparison of different data processing strategies."
#| label: fig-comparison
#| out-width: '90%'

@@

Moving time and moving breath averages yield nearly identical values for V̇O~2max~ over all calculation intervals. This corresponds to the trained athletes reaching respiratory rates around 60 min^-1^ in the final minute of the exercise test (see @fig-rr). Using a zero-phase forward-backward filter (third order, 0.04 Hz cut-off frequency) leads to V̇O~2max~ values 0.4% \[-0.2%; +1.4%\] (median, 10%- and 90%-quantile) higher than the 30-second binned time average approach.

```{r}
#| echo: false
#| fig-cap: "Respiratory rates peak around 60 min^-1^ in the ramp tests. The red segments correspond to the last minute before exhaustion of each individual (n = 72)."
#| fig-scap: "Respiratory rates during the ramp tests."
#| label: fig-rr
#| out-width: '90%'

### 2022-05-31 thesis/thesis.qmd
Include checklist in thesis
Add German abstract to thesis
Improve minor issues in thesis
Finish thesis work

---
title: " "
format: 
  pdf:
    include-in-header: 
      text: \usepackage{setspace}
number-sections: true
top-level-division: section
mainfont: Arial
fontsize: 11pt
urlcolor: blue
linkcolor: black
link-citations: yes
bibliography: ../references.bib
csl: '`r here::here("sports-med-new.csl")`'
geometry: "left = 4cm, right = 2cm, top = 2cm, bottom = 2cm"
params:
  title: 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'
  name: Simon Nolte
  university: German Sport University Cologne
  year: 2022
---

@@

Herewith I affirm in lieu of an oath that I have authored this Master thesis independently and did not use any other sources and tools than indicated. All citations, either direct quotations or passages which were reproduced verbatim or nearby-verbatim from publications, are indicated and the respective references are named. The same is true for tables and figures. I did not submit this piece of work in the same or similar way or in extracts in another assignment.

@@

# Zusammenfassung (German abstract) {-}

@@

@@

@@

>>>>

---
title: " "
format: 
  pdf:
    include-in-header: 
      text: \usepackage{setspace}
    title-meta: 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'
    author-meta: 'Simon Nolte'
    subject: 'Released under a CC-BC-NC-ND license'
number-sections: true
top-level-division: section
mainfont: Arial
fontsize: 11pt
urlcolor: blue
linkcolor: black
link-citations: yes
bibliography: ../references.bib
csl: '`r here::here("sports-med-new.csl")`'
geometry: "left = 4cm, right = 2cm, top = 2cm, bottom = 2cm"
params:
  title: 'Data Processing Strategies to Determine Maximum Oxygen Uptake: A Systematic Scoping Review and Experimental Comparison'
  name: Simon Nolte
  university: German Sport University Cologne
  year: 2022
---

@@

Herewith I affirm in lieu of an oath that I have authored this Bachelor thesis independently and did not use any other sources and tools than indicated. All citations, either direct quotations or passages which were reproduced verbatim or nearby-verbatim from publications, are indicated and the respective references are named. The same is true for tables and figures. I did not submit this piece of work in the same or similar way or in extracts in another assignment.

@@

```{r}
#| child = 'abstract.qmd'
```

@@

\singlespacing

@@

```{r}
#| child = '../transparent_changes.Rmd'
```

@@

![](checklist.pdf)

```{r}
#| echo: false
#| out-extra: "page=2"

knitr::include_graphics("checklist.pdf")
```
